---
title: 'IMT 573: Problem Set 3 - Working With Data II'
author: "Your Name Here"
date: 'Due: Tuesday, October 22, 2019'
output: pdf_document
---

<!-- This syntax can be used to add comments that are ignored during knitting process. -->

##### Collaborators: <!-- BE SURE TO LIST ALL COLLABORATORS HERE! -->

##### Instructions:

Before beginning this assignment, please ensure you have access to R and RStudio; this can be on your own personal computer or on the IMT 573 R Studio Server. 

1. Download the `problemset3.Rmd` file from Canvas or save a copy to your local directory on RStudio Server. Open `problemset3.Rmd` in RStudio and supply your solutions to the assignment by editing `problemset3.Rmd`. 

2. Replace the "Insert Your Name Here" text in the `author:` field with your own full name. Any collaborators must be listed on the top of your assignment. 

3. Be sure to include well-documented (e.g. commented) code chucks, figures, and clearly written text chunk explanations as necessary. Any figures should be clearly labeled and appropriately referenced within the text. Be sure that each visualization adds value to your written explanation; avoid redundancy -- you do not need four different visualizations of the same pattern.

4.  Collaboration on problem sets is fun and useful, and we encourage it, but each student must turn in an individual write-up in their own words as well as code/work that is their own.  Regardless of whether you work with others, what you turn in must be your own work; this includes code and interpretation of results. The names of all collaborators must be listed on each assignment. Do not copy-and-paste from other students' responses or code.

5. All materials and resources that you use (with the exception of lecture slides) must be appropriately referenced within your assignment.  

6. Remember partial credit will be awarded for each question for which a serious attempt at finding an answer has been shown. Students are \emph{strongly} encouraged to attempt each question and to document their reasoning process even if they cannot find the correct answer. If you would like to include R code to show this process, but it does not run without errors, you can do so with the `eval=FALSE` option. (Note: I am also using the `include=FALSE` option here to not include this code in the PDF, but you need to remove this or change it to `TRUE` if you want to include the code chunk.)

```{r example chunk with a bug, eval=FALSE, include=FALSE}
a + b # these object dont' exist 
# if you run this on its own it with give an error
```

7. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit PDF`, rename the knitted PDF file to `ps3_YourLastName_YourFirstName.pdf`, and submit the PDF file on Canvas.

##### Setup: 

In this problem set you will need, at minimum, the following R packages.

```{r Setup, message=FALSE}
# Load standard libraries
library('dplyr')
library('censusr')
library('stringr')
library('lubridate')
library("stringer")
library(ggplot2)
```

#### Problem 1: Joining Census Data to Police Reports

In this problem set, we will be joining disparate sets of data - namely: Seattle police crime data, information on Seattle police beats, and education attainment from the US Census. Our goal is to build a dataset where we can examine questions around crimes in Seattle and the educational attainment of people living in the areas in which the crime occurred; this requires data to be combined from these two individual sources.

As a general rule, be sure to keep copies of the original dataset(s) as you work through cleaning (remember data provenance!).



##### (a) Importing and Inspecting Crime Data

Load the Seattle crime data  from the provided `crime_data.csv` data file. You can find more information on the data here:  \url{https://data.seattle.gov/Public-Safety/Crime-Data/4fs7-3vj5}. This dataset is constantly refreshed online so we will be using the provided csv file for consistency. We will call this dataset the "Crime Dataset." Perform a basic inspection of the Crime Dataset and discuss what you find.


```{r crimedata}
CrimeDataset <- read.csv("crime_data.csv", stringsAsFactors=FALSE)
summary(CrimeDataset)
str(CrimeDataset)
#CrimeDataset$Neighborhood
```

This data represents crimes reported to the Seattle Police Department, each row represents a unique event.
The report number is in numerical format.There is a difference between crime occured and reported date, and these dates have ben loaded as categorical varialbles. The crime sreported have subcategory, Precinct, sector, Beat and Neighborhood as factor variables. Occured and repoted time are in the int format.


##### (b) Looking at Years That Crimes Were Committed

Let's start by looking at the years in which crimes were committed. What is the earliest year in the dataset? Are there any distinct trends with the annual number of crimes committed in the dataset?

Subset the data to only include crimes that were committed after 2011 (remember good practices of data provenance!). Going forward, we will use this data subset.
```{r convertdate}
#We are converting date columns from charecter to date format.

crimedata_withoutna <- na.omit(CrimeDataset)
#We removed all the NA values from CrimeDataSet


crimedata_comitdate <- crimedata_withoutna %>% mutate(CrimeComitDate=
as.Date(crimedata_withoutna$Occurred.Date,"%m/%d/%Y"))

#Occurred.Date column is in char format. We added another column "CrimeComitDate", this column has Occurred.Date values in Dateformat



crime_year_frequency <- crimedata_comitdate %>% group_by(Year=floor_date(CrimeComitDate, "%Y")) %>%
   summarise(Freq_Crime=n()) %>% mutate(Year = substring(Year,1,4)
                                          )

crime_year_frequency

#This crime_year_frequency tells how many crimes were committed every year
```

```{r plotcrimetrends}
#dates$Date <- as.Date(dates$Date)
ggplot(crimedata_comitdate, aes(x=CrimeComitDate)) + geom_histogram(binwidth = 10) +
       scale_x_date(date_labels = "%Y") +
                    
       ylab("CrimeFrequency") + xlab("Year") +
       theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```



```{r years2011andabove}
crimedata_2011above <- subset(crimedata_comitdate, format(as.Date(CrimeComitDate),"%Y")>=2011)
# We applied subset and ctreated another dataframe "crimedata_2011above",
#having crime commit year 2011 and above
```

The earliest year in which crime was committed was 1908.
We can check the histogram we plotted, and also from the "crime_year_frequency" data frame. We notice that hardly a few crimes were committed till the year 1999, however the frequency was gradually increasing. There was some increase from 2000, and then there was a drastic increase in the frequency of committed crimes from 2008.

##### (c) Looking at Frequency of Beats

What is a Police Beat? How frequently are the beats in the Crime Dataset listed? Are there any anomolies with how frequently some of the beats are listed? Are there missing beats?
```{r policebeats}
#crimedata_2011above %>% distinct(Beat)
crimedata_2011above %>% filter(Beat== "K")
crimedata_2011above %>% filter(Beat== "INV")

```
```{r plotting policebeats}
ggplot(crimedata_2011above,aes(x= Beat)) + geom_histogram(stat="count") +
  theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

A police beat ia a small area under a few police personnels in a precinct. 
There may be several police beats in a precinct.
We are only checking the data for crime committed from year 2011 onwards, because it is mentioned in previous question that henceforth we would only check this data.
There are several police beats for each sector, and they are usually made up of two letters. 
Wherever I find police beats made up of one or three letters, their sectors and precincts are unknown.
Also a few rows have missing police beats. 
There are only a few crimes in some police beats while high number of crimes in the others.

##### (d) Importing Police Beat Data and Filtering on Frequency

Load the data on Seattle police beats  provided in `police_beat_and_precinct_centerpoints.csv`. You can find additional information on the data here: (https://data.seattle.gov/Land-Base/Police-Beat-and-Precinct-Centerpoints/4khs-fz35). We will call this dataset the "Beats Dataset."

Does the Crime Dataset include police beats that are not present in the Beats Dataset? If so, how many and with what frequency do they occur? Would you say that these comprise a large number of the observations in the Crime Dataset or are they rather infrequent? Do you think removing them would drastically alter the scope of the Crime Dataset?

Let's remove all instances in the Crime Dataset that have beats which occur fewer than 10 times across the Crime Dataset. Also remove any observations with missing beats. After only keeping years of interest and filtering based on frequency of the beat, how many observations do we now have in the Crime Dataset?

##### (e) Importing and Inspecting Police Beat Data

To join the Beat Dataset to census data, we must have census tract information. Use the `censusr` package to extract the 15-digit census tract for each police beat using the corresponding latitude and longitude. Do this using each of the police beats listed in the Beats Dataset. Do not use a for-loop for this but instead rely on R functions (e.g. the 'apply' family of functions). Add a column to the Beat Dataset that contains the 15-digit census tract for the each beat. (HINT: you may find `censusr`'s `call_geolocator_latlon` function useful)

We will eventually join the Beats Dataset to the Crime Dataset. We could have joined the two and then found the census tracts for each beat. Would there have been a particular advantage/disadvantage to doing this join first and then finding census tracts? If so, what is it? (NOTE: you do not need to write any code to answer this)

##### (f) Extracting FIPS Codes

Once we have the 15-digit census codes, we will break down the code based on information of interest. You can find more information on what these 15 digits represent here: https://transition.fcc.gov/form477/Geo/more_about_census_blocks.pdf.

First, create a column that contains the state code for each beat in the Beats Dataset. Then create a column that contains the county code for each beat. Find the FIPS codes for WA State and King County (the county of Seattle) online. Are the extracted state and county codes what you would expect them to be? Why or why not?

##### (g) Extracting 11-digit Codes

The census data uses an 11-digit code that consists of the state, county, and tract code. It does not include the block code. To join the census data to the Beats Dataset, we must have this code for each of the beats. Extract the 11-digit code for each of the beats in the Beats Dataset. The 11 digits consist of the 2 state digits, 3 county digits, and 6 tract digits. Add a column with the 11-digit code for each beat.

##### (h) Extracting 11-digit Codes From Census

Now, we will examine census data  provided om `census_edu_data.csv`. The data includes counts of education attainment across different census tracts. Note how this data is in a 'wide' format and how it can be converted to a 'long' format. For now, we will work with it as is.

The census data contains a `GEO.id` column. Among other things, this variable encodes the 11-digit code that we had extracted above for each of the police beats. Specifically, when we look at the characters after the characters "US" for values of `GEO.id`, we see encodings for state, county, and tract, which should align with the beats we had above. Extract the 11-digit code from the `GEO.id` column. Add a column to the census data with the 11-digit code for each census observation.

##### (i) Join Datasets

Join the census data with the Beat Dataset using the 11-digit codes as keys. Be sure that you do not lose any of the police beats when doing this join (i.e. your output dataframe should have the same number of rows as the cleaned Beats Dataset - use the correct join). Are there any police beats that do not have any associated census data? If so, how many?

Then, join the Crime Dataset to our joined beat/census data. We can do this using the police beat name. Again, be sure you do not lose any observations from the Crime Dataset. What is the final dimensions of the joined dataset?

Once everything is joined, save the final dataset for future use.