---
title: 'IMT 573: Problem Set 4 - Data Analysis'
author: "Your Name Here"
date: 'Due: Tuesday, October 21, 2019'
output: pdf_document
---

<!-- This syntax can be used to add comments that are ignored during knitting process. -->

##### Collaborators: <!-- BE SURE TO LIST ALL COLLABORATORS HERE! -->

##### Instructions:

Before beginning this assignment, please ensure you have access to R and RStudio; this can be on your own personal computer or on the IMT 573 R Studio Server. 

1. Download the `problemset4.rmd` file from Canvas or save a copy to your local directory on RStudio Server. Open `problemset4.rmd` in RStudio and supply your solutions to the assignment by editing `problemset4.rmd`. 

2. Replace the "Insert Your Name Here" text in the `author:` field with your own full name. Any collaborators must be listed on the top of your assignment. 

3. Be sure to include well-documented (e.g. commented) code chucks, figures, and clearly written text chunk explanations as necessary. Any figures should be clearly labeled and appropriately referenced within the text. Be sure that each visualization adds value to your written explanation; avoid redundancy -- you do no need four different visualizations of the same pattern.

4.  Collaboration on problem sets is fun and useful, and we encourage it, but each student must turn in an individual write-up in their own words as well as code/work that is their own.  Regardless of whether you work with others, what you turn in must be your own work; this includes code and interpretation of results. The names of all collaborators must be listed on each assignment. Do not copy-and-paste from other students' responses or code.

5. All materials and resources that you use (with the exception of lecture slides) must be appropriately referenced within your assignment.  

6. Remember partial credit will be awarded for each question for which a serious attempt at finding an answer has been shown. Students are \emph{strongly} encouraged to attempt each question and to document their reasoning process even if they cannot find the correct answer. If you would like to include R code to show this process, but it does not run withouth errors you can do so with the `eval=FALSE` option.

```{r example chunk with a bug, eval=FALSE, include=FALSE}
a + b # these object dont' exist 
# if you run this on its own it with give an error
```

7. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit PDF`, rename the knitted PDF file to `ps4_ourLastName_YourFirstName.pdf`, and submit the PDF file on Canvas.

##### Setup

In this problem set you will need, at minimum, the following R packages.

```{r Setup, message=FALSE}
# Load standard libraries
library(tidyverse)
library(gridExtra)
```

#### Problem 1: 50 States in the USA

In this problem we will use the `state` dataset, available as part of the R statistical computing platforms. This data is related to the 50 states of the United States of America. Load the data and use it to answer the following questions. 

##### (a) Describe the data and each variable it contains. Tidy the data, preparing it for a data analysis.

##### (b) Suppose you want to explore the relationship between a state's `Murder` rate and other characteristics of the state, for example population, illiteracy rate, and more. Begin by examining the bivariate relationships present in the data. What does your analysis suggest might be important varibles to consider in building a model to explain variation in murder rates?

##### (c) Choose one variable and fit a simple linear regression model, $Y = \beta_1X + \beta_0$, using the `lm()` function in R. Describe your results.

##### (d) Develop a new research question of your own that you can address using the `state` dataset. Clearly state the question you are going to address. Provide at least one visualizations to support your exploration of this question. Discuss what you find.

#### Problem 2: Asking Data Science Questions: Crime and Educational Attainment

In Problem Set 3, you joined data about crimes and educational attainment. Here you will use this new combined dataset to examine questions around crimes in Seattle and the educational attainment of people
living in the areas in which the crime occurred. 

#### (a) Develop a Data Science Question

Develop your own question to address in this analysis. Your question should be specific and measurable, and it should be able to be addressed through a basic analysis of the crime dataset you compiled in Problem Set 3. 

#### (b) Describe and Summarize

Briefly summarize the dataset, describing what data exists and its basic properties. Comment on any issues that need to be resolved before you can proceed with your analysis. 

#### (c) Data Analysis

Use the dataset to provide empirical evidence that addressed your question from part (a). Discuss your results. Provide at least one visualization to support your narrative. 

#### (d) Reflect and Question

Comment the questions (and answers) in this analysis.  Were you able to answer all of these questions?  Are all questions well defined?  Is the data good enough to answer all these?

#### Problem 3: Sampling with and without Replacement

In the following situations assume that half of the specified population wears hats every day and the other half does not wear hats. 

##### (a) Suppose you're sampling from a room with 10 people. What is the probability of sampling two hat-wearing people in a row when sampling with replacement? What is the probability when sampling without replacement?

##### (b)  Now suppose you're sampling from a stadium with 10,000 people. What is the probability of sampling two hat wearers in a row when sampling with replacement? What is the probability when sampling without replacement?

##### (c) We often treat individuals who are sampled from a large population as independent. Using your findings from parts (a) and (b), explain whether or not this assumption is reasonable.
