---
title: 'IMT 573: Problem Set 1 - Exploring Data'
author: "Divya Gaurav Tripathi"
date: 'Due: Tuesday, October 08, 2019'
output: pdf_document
---

<!-- This syntax can be used to add comments that are ignored during knitting process. -->

##### Collaborators: <!-- BE SURE TO LIST ALL COLLABORATORS HERE! --> N/A

##### Instructions:

Before beginning this assignment, please ensure you have access to R and RStudio; this can be on your own personal computer or on the IMT 573 R Studio Server. 

1. Download the `problemset1.Rmd` file from Canvas or save a copy to your local directory on RStudio Server. Open `problemset1.Rmd` in RStudio and supply your solutions to the assignment by editing `problemset1.Rmd`. 

2. Replace the "Insert Your Name Here" text in the `author:` field with your own full name. Any collaborators must be listed on the top of your assignment. 

3. Be sure to include well-documented (e.g. commented) code chucks, figures, and clearly written text chunk explanations as necessary. Any figures should be clearly labeled and appropriately referenced within the text. Be sure that each visualization adds value to your written explanation; avoid redundancy -- you do no need four different visualizations of the same pattern.

4.  Collaboration on problem sets is fun and useful, and we encourage it, but each student must turn in an individual write-up in their own words as well as code/work that is their own.  Regardless of whether you work with others, what you turn in must be your own work; this includes code and interpretation of results. The names of all collaborators must be listed on each assignment. Do not copy-and-paste from other students' responses or code.

5. All materials and resources that you use (with the exception of lecture slides) must be appropriately referenced within your assignment.  

6. Remember partial credit will be awarded for each question for which a serious attempt at finding an answer has been shown. Students are \emph{strongly} encouraged to attempt each question and to document their reasoning process even if they cannot find the correct answer. If you would like to include R code to show this process, but it does not run withouth errors you can do so with the `eval=FALSE` option as follows:

```{r example chunk with a bug, eval=FALSE}
a + b # these object dont' exist 
# if you run this on its own it with give an error
```

7. When you have completed the assignment and have **checked** that your code both runs in the Console and knits correctly when you click `Knit PDF`, rename the knitted PDF file to `Yps1_ourLastName_YourFirstName.pdf`, and submit the PDF file on Canvas.

##### Setup: 

In this problem set you will need, at minimum, the following R packages.

```{r Setup, message=FALSE}
# Load standard libraries
library(tidyverse)
library(nycflights13)
library(ggplot2)
```

#### Problem 1: Exploring the NYC Flights Data

In this problem set we will use the data on all flights that departed NYC (i.e. JFK, LGA or EWR) in 2013. You can find this data in the `nycflights13` R package. 

```{r}
# Load the nycflights13 library which includes data on all
# lights departing NYC
data(flights)
# Note the data itself is called flights, we will make it into a local df
# for readability
flights <- tbl_df(flights)
# Look at the help file for information about the data
?flights
flights
summary(flights)
#This would tell me the structure of variables in this data frame.
str(flights)
#I want to find out if the data is only about 2013
flights %>% distinct(year)
#I want to find out various destinations of these flights
flights %>% distinct(dest)
flights %>% distinct(carrier)
```

##### (a) Importing and Inspecting Data

Load the data and describe in a short paragraph how the data was collected and what each variable represents. Perform a basic inspection of the data and discuss what you find.

The data has been loaded in the above code chunk. This data is about the flights departing from various airports of NewYork (i.e. JFK, LGA or EWR) in the year 2013.There are 336776 observations of 19 variables.The variables are of different types such as int, num, char and POSIXct date.
The data tells us about various aspects of the flights going out of NewYork in 2013. The Bureau of Transportation Statistics had collected this data. This is an observationla study for only the year 2013 and only flights departing from varioius airports of NewYork. This data cannot be interpretted as a representative of general population (i.e. for the flights from all the US airports every year).
As there was no random assignment, no causation can be infered.
The meaning of various variables are as following: carrier- abbreviation for flight carrier, tailnum- plane tail number, origin- originating airport, dest- destination airport,air_time- number of minutes flight spent in air, distance- distance between airports in miles, time_hour- scheduled date and hour of flight, year,month,day- Date of flight departure,
dep_time,arr_time- actual departure and arrival times,sched_dep_time,sched_arr_time- scheduled departure and arrival times,dep_delay,arr_delay-Departure and arrival delays, hour,minute-Time of scheduled departure broken into hour and minutes.

##### (b) Formulating Questions

Consider the NYC flights data. Formulate two motivating questions you want to explore using this data. Describe why these questions are interesting and how you might go about answering them.

Q1. Out of all the flights departing from NewYork and arriving in Seattle, what was the mean arrival delay for various carriers? I live in Seattle, so this data woul help me in choosing flights.I would try to find the mean from the given data set and plot its graph for different carriers.The Na values have to be removed and data has to be cleaned which might consist adding of new columns and creating different data frames.

Q2. What was the mean time spent by some selected carriers ('DL','B6','UA','AS','AA') in air during various months of 2013? If I am an analyst who wants to study how much business various airlines are doing, I should find out this. I would try to find the mean for air time for these carriers for each months. I would have to group by carrier and montgh, and then plot it. 

##### (c) Exploring Data

For each of the questions you proposed in Problem 1b, perform an exploratory data analysis designed to address the question. At a minimum, you should produce two visualizations (graphics or tables) related to each question. Be sure to describe what the visuals show and how they speak to your question of interest. 


We would explore our first question which was as following:

Q1. Out of all the flights departing from NewYork and arriving in Seattle, what was the mean arrival delay for various carriers? 

```{r q1_dataframe}
# I would ctreate a new data frame for question one for the flights having Sea (Seattle) as their destination. I would filter out the NA values

q1_dataframe <- flights %>% filter(!is.na(flights$carrier)) %>% filter(dest == 'SEA') %>% select(carrier,arr_delay) %>% filter(arr_delay>0) 

# I am not interested in flights arriving early, so I have filtered for arrival delay >0

q1_df_grouped <- q1_dataframe %>% group_by(carrier) %>% mutate(meandelay = mean(arr_delay))



#I have created another data frame, grouped by carrier and added another column for mean arrival delay of each of carrier
```

```{r bargraph_Q1}
ggplot(data = q1_df_grouped,  aes(x= carrier, y = meandelay)) + geom_bar(stat = "identity",fill = "#FF1111") + labs(x = "Carrier Name", y= "Mean Delay in Mins", title = "Delay for Carrier") + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())#scale_fill_grey(start = 0, end = 15000)
```
The above plot is about mean arrival delay for carriers that arrive in Seattle. The above plot clearly tells that the mean delay in minutes is least for AA carrier and highest for UA carrier. This might indicate UA carrier often does not run on schedule.

```{r boxplot_Q1}
boxplot(data= q1_dataframe, arr_delay ~ carrier )

```

This box plot shows many outliers for the DL carrier, suggesting some of its flights wee extremely delayed in arrival.



We would now explore our second question which was as following:

Q2. What was the total timr spent by various carriers in air? If I am an analyst who wants to study how much business various airlines are doing, I should find out this.



```{r df_Q2}
#We would make a data frame for this question.We are only analysing data for the following carriers: DL,B6,UA, AS, AA. We would also remove the NA values of month and air_time.

df_Q2 <- flights %>% select(carrier, air_time, month) %>% filter(carrier %in% c('DL','B6','UA','AS','AA')) %>% filter(!is.na(month)) %>% filter(!is.na(air_time))



month_names_func <- Vectorize(function(n) c("Jan", "Feb", "Mar", 
                                         "Apr", "May", "Jun", "Jul", 
                                         "Aug", "Sep", "Oct",
                                         "Nov", "Dec")[n])

df_monthname <- df_Q2 %>% mutate(month_name = month_names_func(month))

#We created a function month_names_func, and used it while adding another column(month_name); which would display the name of month instead of its digit. We stored the result in another data frame named df_monthname.

df_Q2_grouped <- df_monthname  %>% group_by(carrier,month,month_name) %>%  mutate(mean_air_time = mean(air_time)) %>% distinct(carrier,month_name,mean_air_time) 

#f_factor <- df_Q2_grouped %>% factor(month_name,levels = ("Jan", "Feb", "Mar","Apr", "May", "Jun", "Jul","Aug", "Sep", "Oct","Nov", "Dec"))

#df_Q2_grouped %>% distinct(mean_air_time)

# We grouped with carrier and month name and found out the mean air time for each carrier in every month.

```

```{r Q2_plot, fig.height = 10, fig.width =12}
ggplot(data=df_Q2_grouped, aes(x= month_name,y=mean_air_time, color = month_name)) + geom_bar(stat = "identity") + labs(x = "Month", y= "Mean Air Time", title = "Air Time In Month For Various Carriers")+ facet_wrap(~df_Q2_grouped$carrier) + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

```

The above plots show the mean air time for some carriers ('DL','B6','UA','AS','AA') during different months of 2013. From the above plots about these five airlines, we find that the mean air time was highest for AS carrier throughout the year. There is not much variation in the distribution of mean air times of different carriers in various months. This may indicate that AS carrier was doing the most business in the year 2013.

##### (d) Challenge Your Results

After completing the exploratory analyses from Problem 1c, do you have any concerns about your findings? How well defined was your original question? Do you still believe this question can be answered using this dataset? Comment on any ethical and/or privacy concerns you have with your analysis. 

My finding of for question 1 shows that UA has the most delayed flights from NewYork to Seattle. However, this might not reflect the actual performance of the airline; it may be that in bad weather other airlines prefer to cancel their flights but this airline does not. It might also be that this airline performs mostly during high traffic days, which might cause delay. Sucg external factors might cause bias while assessing the performance of airlines.
I also found out that the DL carrier had a few flights that were extremely delayed compared to other carriers. This might also be due to external factors and should not be used to infer its performance.

My finding for question 2 said that AS carrier  doing the most business based on its mean time spent in air. However, this might not mean it is making the most profit, as this airline might be owning a larger fleet of airplanes which would also require greater investment. The other airlines might be spending less time in air but may be earning higher profit becuase of ticket cost and better services, these factors were not a part of this study.

In my opinion, this study reveals no private data about stakeholder, and is only a collection of publicly available data, so privacy is not a concern. However, there are various aspects of airlines which cannot be fully quantified (like weather condition, traffic condition, customer satisfaction, service quality of airlines) and were not a part of this study. If we try to assess the performance of airlines only based on its schedule, travel distance or flight time; we might get biased results. So ethically speaking, this study should not be used to infer performance of airlines.

References:
Bureau of Transportation Statistics: https://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236 


